Developing a Bubble Chamber Particle Discriminator Using Semi-Supervised Learning

### 2019-02-26 documentation/report.tex
Resized charts to fit text
Made everything much clearer

@@

\begin{center}
    \begin{LARGE}
        Improving Particle Classification in WIMP Dark Matter Detection Experiments Using Neural Networks
    \end{LARGE}

@@

In DEAP-3600, in a simulated environment, a conventional classifier removes 99.6\% of alpha background radiation, while also (undesirably) removing 91.0\% of WIMP events.

@@

\begin{figure}[ht]
    \centering
    \subfloat{\fbox{\includegraphics[width=0.95\textwidth]{audio_cropped}}}
    \qquad
    \subfloat{\fbox{\includegraphics[width=0.45\textwidth]{image_grid}}}
    \caption{Examples of the audio waveform and Fourier transform (above) and the images captured by cameras (below).}
\end{figure}

@@

\begin{enumerate}
    \item Iterative Cluster Nucleation is inspired by unsupervised "clustering" algorithms, and takes advantage of the ``confidence'' of the neural network's predictions (that is, how close the predictions are to 0 or 1).

@@

    \item I developed and applied a Mercator-like cylindrical projection, by mapping each row of nodes in the sphere to a row in a rectangular image. Cubic spline interpolation was used to map latitude lines on the sphere (which vary in length) to equal-length rows on the rectangular image.

@@

\begin{figure}[ht]
    \centering
    \subfloat{\fbox{\includegraphics[width=0.45\textwidth]{map_projection}}}
    \qquad
    \subfloat{\fbox{\includegraphics[width=0.45\textwidth]{topological}}}
    \caption{Diagrams of the map projection (left) and topological CNN (right) systems.}
\end{figure}

@@

As seen in Figure \ref{pico_final_results}, the banded FFT input format was the most effective of all of the supervised learning models (including the previous study), reaching 97.3\% accuracy on validation data. The image data was shown to contain no meaningful information; it permitted only 63.0\% accuracy.

Semi-supervised learning produced better accuracy than supervised learning. Gravitational differentation was the best overall model, reaching 99.7\% accuracy on validation data and 98.3\% on test data. This confirms my hypothesis; it is indeed beneficial to train on the neural network's most confident predictions.

\begin{figure}[ht]
    \centering
    \fbox{\includegraphics[width=0.9\textwidth]{pico_final_results}}
    \caption{\label{pico_final_results} Validation accuracy, precision, recall, and loss for each technique in PICO-60. Error bars represent the 8th and 92nd percentiles.}
\end{figure}

@@

Both the dense neural network and topological CNN produced a higher rate of false positives than the conventional classifier. However, as seen in Figure \ref{deap_final_results}, the cylindrical projection method improved the false positive rate from 91.0\% to 75.7\% (a 63.0\% reduction)!

When tested on real-world data, the cylindrical projection method removed 73.9\% of nuclear recoil (neutron calibration) events (in line with the simulation).

However, only 91.2\% of background radiation events were removed. This indicates either that the simulation is not an accurate representation of the real-world detector, or (less probably) that WIMPs are present in the background data.

\begin{figure}[ht]
    \centering
    \fbox{\includegraphics[width=0.55\textwidth]{deap_final_results}}
    \caption{\label{deap_final_results} Background and WIMP removal rates for the highest-accuracy hyperparameter configuration of each technique in DEAP-3600.}
\end{figure}

>>>>

\titleformat*{\section}{\large\bfseries}
\titleformat*{\subsection}{\normalsize\bfseries}
\titleformat*{\subsubsection}{\normalsize\bfseries}

@@

\begin{center}
    \begin{Large}
        Improving Particle Classification in WIMP Dark Matter Detection Experiments Using Neural Networks
    \end{Large}

@@

In DEAP-3600, a simulation was developed prior to my work, because it is impractical to collect large amounts of calibration data. A conventional classifier removes 99.6\% of alpha background radiation, while also (undesirably) removing 91.0\% of simulated WIMP events.

@@

\begin{figure}[ht]
    \centering
    \subfloat{\fbox{\includegraphics[width=0.8\textwidth]{audio_cropped}}}
    \qquad
    \subfloat{\fbox{\includegraphics[width=0.45\textwidth]{image_grid}}}
    \caption{Examples of the audio waveform and Fourier transform (above) and the images captured by cameras (below).}
\end{figure}

@@

\begin{enumerate}
    \item Iterative Cluster Nucleation is inspired by unsupervised clustering algorithms, and takes advantage of the ``confidence'' of the neural network's predictions (that is, how close the predictions are to 0 or 1).

@@

    \item I developed and applied a Mercator-like cylindrical projection, which maps a sphere onto a rectangle. Cubic spline interpolation was used to map latitude lines on the sphere (which vary in length) to equal-length rows on the rectangular image.

@@

\begin{figure}[ht]
    \centering
    \fbox{\includegraphics[width=0.45\textwidth]{map_projection}}
    \qquad
    \fbox{\includegraphics[width=0.45\textwidth]{topological}}
    \caption{Diagrams of the cylindrical projection (left) and topological CNN (right) systems.}
\end{figure}

@@

As seen in Figure \ref{pico_final_results}, the banded FFT format was the most effective of the supervised learning models (including the previous study), reaching 97.3\% accuracy on validation data. The image data was shown to contain no meaningful information, permitting only 63.0\% accuracy.

Semi-supervised learning produced better accuracy than supervised learning. Gravitational differentiation was the best overall model, reaching 99.7\% accuracy on validation data and 98.3\% on test data. This confirms my hypothesis; it is indeed beneficial to train on the neural network's most confident predictions.

\begin{figure}[ht]
    \centering
    \fbox{\includegraphics[width=0.85\textwidth]{pico_final_results}}
    \caption{\label{pico_final_results} Validation accuracy, precision, recall, and loss for each technique in PICO-60. Error bars represent the 8th and 92nd percentiles.}
\end{figure}

@@

Both the dense neural network and topological CNN produced a higher rate of false positives than the conventional classifier, meaning that they misclassified more WIMPs as alpha particles. However, as seen in Figure \ref{deap_final_results}, the cylindrical projection method \textit{reduced} the false positive rate from 91.0\% to 75.7\%.

This successful network model was tested on the limited amount of available real-world data. It removed 73.9\% of WIMP-like (neutron calibration) events, which is in line with the 75.7\% obtained on simulated data.

However, only 91.2\% of background events (expected to be neck alphas) were removed, quite far off the 99.6\% reached on simulated data. This indicates one of three possibilities:

\begin{enumerate}
    \item The simulation is not an accurate representation of the real-world detector (some physical dynamic has not been simulated), and thus the network did not generalize.
    \item There is WIMP dark matter present in the set of background events.
    \item There is an unanticipated form of background radiation present.
\end{enumerate}

\begin{figure}[ht]
    \centering
    \fbox{\includegraphics[width=0.5\textwidth]{deap_final_results}}
    \qquad
    \begin{tabular}[b]{|l|l|l|}
        \hline
        Method & Backgnd. & WIMP \\
        \hline
        Conventional & 99.6\% & 91.0\% \\
        \hline
        Dense NN & 100\% & 100\% \\
        \hline
        Cylindrical & 99.9\% & 76.7\% \\
        \hline
        Topological & 99.9\% & 93.0\% \\
        \hline
        Cyl Sim Test & 99.6\% & 75.7\% \\
        \hline
        Cyl Real Test & 91.2\% & 73.9\% \\
        \hline
    \end{tabular}
    \caption{\label{deap_final_results} Background and WIMP removal rates for the highest-accuracy hyperparameter configuration of each technique in DEAP-3600.}
\end{figure}
